---
title: "Disease-group comparison with spatialCooccur"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Disease-group comparison with spatialCooccur}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  dpi = 90
)
```

This vignette walks through the disease- / condition-group comparison
workflow in **spatialCooccur**. The pattern is:

1. Collect per-sample spatial transcriptomics data (one image per sample,
   or multiple images grouped by patient).
2. Compute a spatial co-occurrence score *per sample* with one of the
   `*_per_sample()` helpers.
3. Test for differences between groups (e.g. disease vs. control) with
   `compare_groups()`.
4. Visualize with `plot_group_delta_heatmap()`, `plot_pair_boxplot()`, or
   `plot_volcano_groups()`.

```{r load}
library(spatialCooccur)
```

# Simulate a two-group dataset

## How `generate_sim()` builds one sample

Before generating the full disease/control dataset, it helps to look at
what a single sample produced by `generate_sim()` actually contains.

A sample is a flat `data.frame` of cells with `x`, `y`, and `cell_type`
columns. With `test_type = "distribute"`:

1. All cells are first dropped uniformly at random in a `max_loc x
   max_loc` box.
2. Each cell is assigned one of `n_types` cell types.
3. A fraction `close_ratio` of the `cell_type_2` cells is then *moved*
   next to a randomly chosen `cell_type_1` cell, at distance
   `distance_param` (with small Gaussian jitter).

So `close_ratio` controls the strength of the planted 1 ↔ 2 spatial
interaction. `distance_param` controls how *close* the pairing is.

```{r sim_anatomy, fig.width = 9, fig.height = 3, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)

# A small parameter sweep on close_ratio, all other params fixed.
ratios  <- c(0.0, 0.3, 0.7, 1.0)
df_list <- lapply(ratios, function(r) {
  d <- generate_sim(close_ratio    = r,
                    n_types        = 5,
                    n_cells        = 400,
                    max_loc        = 300,
                    test_type      = "distribute",
                    distance_param = 15,
                    seed           = 7)
  d$close_ratio <- sprintf("close_ratio = %.1f", r)
  d$.target     <- ifelse(as.character(d$cell_type) %in%
                            c("cell_type_1", "cell_type_2"),
                          as.character(d$cell_type), "other")
  d
})
df_sweep <- do.call(rbind, df_list)
df_sweep$close_ratio <- factor(df_sweep$close_ratio,
                               levels = sprintf("close_ratio = %.1f", ratios))

ggplot(df_sweep, aes(x = x, y = y)) +
  geom_point(data = subset(df_sweep, .target == "other"),
             color = "grey85", size = 0.35) +
  geom_point(data = subset(df_sweep, .target != "other"),
             aes(color = .target), size = 0.9) +
  facet_wrap(~ close_ratio, nrow = 1) +
  coord_equal() +
  scale_color_manual(values = c(cell_type_1 = "#E41A1C", cell_type_2 = "#377EB8")) +
  labs(title = "Effect of close_ratio on one simulated sample",
       subtitle = "grey = other cell types, red = cell_type_1, blue = cell_type_2",
       x = NULL, y = NULL, color = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank())
```

As `close_ratio` grows from 0 to 1 you can see the red and blue points
visibly clump into pairs while the grey background stays uniform. That
clumping is exactly the signal `nhood_enrichment` should pick up.

To make `distance_param` concrete, zoom into one sample and draw a
segment from each red seed to its paired blue partner:

```{r sim_distance_param, fig.width = 6, fig.height = 5, eval = requireNamespace("ggplot2", quietly = TRUE)}
one <- generate_sim(close_ratio    = 1.0,
                    n_types        = 5,
                    n_cells        = 300,
                    max_loc        = 300,
                    test_type      = "distribute",
                    distance_param = 25,
                    seed           = 42)

# Reconstruct the seed-partner pairs deterministically by re-running the
# seeding logic of generate_sim() so we can draw the connecting segments.
set.seed(42)
.x  <- runif(300, 0, 300); .y <- runif(300, 0, 300)
.ct <- sample(paste0("cell_type_", 1:5), 300, replace = TRUE)
i1  <- which(.ct == "cell_type_1")
i2  <- which(.ct == "cell_type_2")
n_pairs <- round(length(i2) * 1.0)
ic1 <- sample(i1, n_pairs, replace = TRUE)
ic2 <- sample(i2, n_pairs)

# Use the moved coordinates of cell_type_2 from the simulated sample,
# matched back by index.
segs <- data.frame(
  x = one$x[ic1], y = one$y[ic1],
  xend = one$x[ic2], yend = one$y[ic2]
)
# Keep only nearby pairs for clarity in the zoom
zoom <- segs[segs$x > 50 & segs$x < 230 & segs$y > 50 & segs$y < 230, ]

ggplot() +
  geom_point(data = one, aes(x = x, y = y),
             color = "grey85", size = 0.4) +
  geom_segment(data = zoom,
               aes(x = x, y = y, xend = xend, yend = yend),
               color = "grey40", linewidth = 0.3) +
  geom_point(data = subset(one, cell_type == "cell_type_1"),
             aes(x = x, y = y), color = "#E41A1C", size = 1.2) +
  geom_point(data = subset(one, cell_type == "cell_type_2"),
             aes(x = x, y = y), color = "#377EB8", size = 1.2) +
  coord_equal() +
  labs(title = "Each blue (cell_type_2) sits ≈ distance_param away from a red seed",
       subtitle = "lines connect each seed (cell_type_1) to its paired partner (cell_type_2)",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

`generate_sim_groups()` then calls `generate_sim()` once per sample,
using a group-specific `close_ratio` (plus a small per-sample noise), so
the result is `n_samples_per_group x length(group_close_ratio)` samples
sharing the same coordinate space but differing in how strongly the 1 ↔ 2
interaction is planted.

```{r simulate}
set.seed(1)
df_groups <- generate_sim_groups(
  n_samples_per_group = 4,
  group_close_ratio = list(disease = 0.85, control = 0.15),
  n_types = 5,
  n_cells = 400,
  test_type = "distribute",
  distance_param = 15,
  between_sample_noise = 0.03,
  seed = 1
)
head(df_groups)
table(df_groups$sample_id, df_groups$group)
```

`df_groups` is a long-format data.frame with columns `x`, `y`,
`cell_type`, `sample_id`, `group`, `patient`. The same shape is what the
`*_per_sample()` helpers accept directly; for Seurat input they read the
equivalent fields from `meta.data` and `@images`.

## Spatial map of the simulated samples

Before any statistics, it helps to see the planted structure: with
`test_type = "distribute"` and `group_close_ratio = list(disease = 0.85,
control = 0.15)`, the disease samples should show clear local pairing of
`cell_type_1` (the seed type) and `cell_type_2` (the partner placed at
`distance_param` from each seed), while control samples should look
nearly random.

```{r spatial_map, fig.width = 8, fig.height = 4.5, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)

# Order panels so disease and control sit on contiguous rows
df_groups$sample_id <- factor(
  df_groups$sample_id,
  levels = unique(df_groups$sample_id[order(df_groups$group, df_groups$sample_id)])
)

ggplot(df_groups, aes(x = x, y = y, color = cell_type)) +
  geom_point(size = 0.6, alpha = 0.8) +
  facet_wrap(~ group + sample_id, nrow = 2) +
  coord_equal() +
  scale_color_manual(values = manual_colors[seq_len(nlevels(df_groups$cell_type))]) +
  guides(color = guide_legend(override.aes = list(size = 2.5))) +
  labs(title = "Simulated samples by disease group",
       x = NULL, y = NULL, color = "cell type") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank())
```

To highlight the planted interaction, restrict to the two seeded cell
types and overlay them on a grey background of all other cells:

```{r spatial_map_focus, fig.width = 8, fig.height = 4.5, eval = requireNamespace("ggplot2", quietly = TRUE)}
target <- c("cell_type_1", "cell_type_2")
df_groups$.target <- ifelse(as.character(df_groups$cell_type) %in% target,
                            as.character(df_groups$cell_type), "other")

ggplot(df_groups, aes(x = x, y = y)) +
  geom_point(data = subset(df_groups, .target == "other"),
             color = "grey85", size = 0.4) +
  geom_point(data = subset(df_groups, .target != "other"),
             aes(color = .target), size = 0.9) +
  facet_wrap(~ group + sample_id, nrow = 2) +
  coord_equal() +
  scale_color_manual(values = c(cell_type_1 = "#E41A1C", cell_type_2 = "#377EB8")) +
  labs(title = "Planted interaction: cell_type_1 vs cell_type_2",
       subtitle = "disease samples should show tight 1–2 pairing; control samples should not",
       x = NULL, y = NULL, color = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank())

df_groups$.target <- NULL
```

# Per-sample neighborhood enrichment

```{r per_sample, message = FALSE}
per_sample <- nhood_enrichment_per_sample(
  df_groups,
  sample_key   = "sample_id",
  group_key    = "group",
  cluster_key  = "cell_type",
  patient_key  = "patient",
  neighbors.k  = 15,
  n_perms      = 100,
  n_jobs       = 1,
  seed         = 1
)
head(per_sample)
```

Each row is `sample_id x cluster_i x cluster_j`. The factor levels of
`cell_type` are harmonized across samples so that the `cluster_i`,
`cluster_j` columns are directly comparable.

For the radius-based ratio, the local interaction score, and the
interaction-spot count, use the parallel helpers:

```{r other_per_sample, eval = FALSE}
cooccur_ratio_per_sample(
  df_groups, sample_key = "sample_id", group_key = "group",
  cluster_key = "cell_type", patient_key = "patient",
  radius = 30, k = 15
)

cooccur_local_per_sample(
  df_groups, sample_key = "sample_id", group_key = "group",
  cluster_key = "cell_type",
  cluster_x = "cell_type_1", cluster_y = "cell_type_2",
  patient_key = "patient",
  neighbors.k = 15, radius = 30
)
```

# Group comparison

`compare_groups()` runs a per-cluster-pair test between two groups. Set
`ref_group` to make the sign of `effect` interpretable
(`effect = mean(test) - mean(reference)`).

## Algorithm

Let $S$ be the set of samples, $G_s \in \{g_\mathrm{ref}, g_\mathrm{test}\}$
the group of sample $s$, and $V_{s,i,j}$ the per-sample score (e.g. the
z-score from `nhood_enrichment_per_sample()`) for the ordered cluster
pair $(i, j)$. `compare_groups()` iterates over every unique
$(i, j)$ and runs the chosen two-group test on the vector
$\{V_{s,i,j}\}_{s \in S}$ split by $G_s$.

For each pair it reports

$$
\hat{\mu}_g \;=\; \frac{1}{|S_g|}\sum_{s \in S_g} V_{s,i,j},
\qquad
\widehat{\text{effect}}_{i,j} \;=\; \hat{\mu}_{g_\mathrm{test}} - \hat{\mu}_{g_\mathrm{ref}},
$$

where $S_g = \{s : G_s = g\}$.

After all pairs are tested, raw $p$-values $p_{i,j}$ are corrected for
multiple testing across pairs:

$$
\tilde{p}_{i,j} \;=\; \mathrm{adjust}\bigl(\{p_{i,j}\};\, \text{method}\bigr),
$$

defaulting to the Benjamini–Hochberg FDR procedure
($\tilde{p}_{(k)} = \min_{m \ge k}\bigl\{\frac{M}{m} p_{(m)}\bigr\}$, with
$M$ the number of pairs and $p_{(k)}$ the $k$-th smallest $p$-value).

The four `method` choices differ only in how each per-pair $p_{i,j}$ is
computed.

### `method = "wilcox"` — Wilcoxon rank-sum

Let $V^{(\mathrm{ref})}$ and $V^{(\mathrm{test})}$ denote the per-sample
scores in each group (sizes $n_1, n_2$), and let $R_k$ be the rank of
$V^{(\mathrm{test})}_k$ when all $n_1 + n_2$ values are pooled and ranked.
The Mann–Whitney $U$ statistic is

$$
U \;=\; \sum_{k=1}^{n_2} R_k \;-\; \frac{n_2(n_2+1)}{2}.
$$

Under the null ($V^{(\mathrm{ref})}$ and $V^{(\mathrm{test})}$ share a
distribution),

$$
\mathbb{E}[U] = \frac{n_1 n_2}{2}, \quad
\mathrm{Var}(U) = \frac{n_1 n_2 (n_1 + n_2 + 1)}{12},
$$

and the two-sided $p$-value comes from the normal approximation in
`stats::wilcox.test(..., exact = FALSE)`.

This test is non-parametric and robust, which makes it the safest
default when you have a handful of samples per group.

### `method = "t"` — Welch's two-sample $t$

Same per-pair split as above. Let $\bar V_g$, $s_g^2$ be the sample mean
and unbiased variance in each group. Welch's statistic is

$$
t \;=\; \frac{\bar V^{(\mathrm{test})} - \bar V^{(\mathrm{ref})}}
              {\sqrt{\dfrac{s_1^2}{n_1} + \dfrac{s_2^2}{n_2}}}
$$

with Welch–Satterthwaite degrees of freedom

$$
\nu \;=\; \frac{\left(\dfrac{s_1^2}{n_1} + \dfrac{s_2^2}{n_2}\right)^2}
                {\dfrac{(s_1^2/n_1)^2}{n_1 - 1} + \dfrac{(s_2^2/n_2)^2}{n_2 - 1}}.
$$

Use this if you trust approximate normality of the per-sample scores.

### `method = "lmm"` — linear mixed model

When you have multiple images per patient, you should not treat them as
independent observations. For a given pair $(i, j)$, write the score of
image $k$ from patient $p$ as

$$
V_{p,k,i,j} \;=\; \beta_0 \;+\; \beta_1 \, \mathbb{1}\{G_{p,k} = g_\mathrm{test}\}
                 \;+\; u_p \;+\; \varepsilon_{p,k},
$$

with random patient intercept $u_p \sim \mathcal{N}(0, \sigma_u^2)$ and
residuals $\varepsilon_{p,k} \sim \mathcal{N}(0, \sigma_\varepsilon^2)$.

`compare_groups()` fits this with `lme4::lmer(value ~ group + (1 |
patient))`, then tests $H_0\!: \beta_1 = 0$ with a Wald-z approximation
using the $t$-value from `summary(fit)$coefficients`:

$$
z \;=\; \frac{\hat\beta_1}{\widehat{\mathrm{SE}}(\hat\beta_1)},
\qquad
p \;=\; 2\,\Phi(-|z|).
$$

If each patient contributes only a single observation the LMM collapses
to OLS and `compare_groups()` falls back to `lm()`. The Wald-z $p$-value
is approximate; for stricter inference fit with `lmerTest` outside the
package.

### `method = "perm"` — group-label permutation

Let $\Delta = \hat\mu_{g_\mathrm{test}} - \hat\mu_{g_\mathrm{ref}}$ be
the observed mean difference for pair $(i, j)$. We then shuffle the
group labels $B$ times (`n_perms`) to build a null distribution.

Without `patient_key`, every permutation reassigns sample-level labels
independently:

$$
\Delta^{(b)} \;=\; \hat\mu^{(b)}_{g_\mathrm{test}} - \hat\mu^{(b)}_{g_\mathrm{ref}},
\quad b = 1, \dots, B.
$$

With `patient_key`, labels are shuffled at the patient level (each
patient keeps the same label across all of its images), which preserves
within-patient correlation.

The two-sided $p$-value uses the standard $+1$-smoothed estimator

$$
p \;=\; \frac{1 \;+\; \#\{b : |\Delta^{(b)}| \ge |\Delta|\}}{1 \;+\; B},
$$

avoiding $p = 0$ in finite-$B$ samples. This is the preferred method
when the per-sample score distribution is far from normal or when you
need patient-level blocking.

### Multiple-testing adjustment

All $p$-values are passed through `stats::p.adjust(..., method =
adjust)`. The default `"BH"` controls the false discovery rate across the
$M$ tested cluster pairs at the chosen level $\alpha$ by

$$
k^{*} \;=\; \max\Bigl\{k : p_{(k)} \le \frac{k}{M}\,\alpha\Bigr\},
$$

and rejecting all pairs with $p_{(k)} \le p_{(k^*)}$. The returned
`padj` column holds the BH-adjusted values directly, so a downstream
$\tilde p_{i,j} < \alpha$ check gives exactly the BH decision.

```{r wilcox}
cmp_wilcox <- compare_groups(
  per_sample,
  value     = "zscore",
  method    = "wilcox",
  adjust    = "BH",
  ref_group = "control"
)
head(cmp_wilcox)
```

When you have multiple images per patient, prefer a mixed model (random
intercept on patient) or a patient-blocked permutation:

```{r lmm, eval = requireNamespace("lme4", quietly = TRUE)}
cmp_lmm <- compare_groups(
  per_sample,
  value       = "zscore",
  method      = "lmm",
  patient_key = "patient",
  ref_group   = "control"
)
head(cmp_lmm)
```

```{r perm}
cmp_perm <- compare_groups(
  per_sample,
  value       = "zscore",
  method      = "perm",
  patient_key = "patient",
  n_perms     = 500,
  ref_group   = "control",
  seed        = 1
)
head(cmp_perm)
```

# Visualization

```{r plots, eval = requireNamespace("ggplot2", quietly = TRUE), fig.show = "hold"}
plot_group_delta_heatmap(cmp_wilcox)
plot_volcano_groups(cmp_wilcox, label_top = 5)
plot_pair_boxplot(
  per_sample,
  value     = "zscore",
  pairs     = data.frame(cluster_i = "cell_type_1",
                         cluster_j = "cell_type_2"),
  add_p     = TRUE,
  ref_group = "control"
)
```

# Patient-level aggregation

If you have multiple images per patient and want one observation per
patient, pass `unit = "patient"` to the `*_per_sample()` helpers. Each
image-level result is then averaged within patient before group
comparison:

```{r patient_level, eval = FALSE}
per_patient <- nhood_enrichment_per_sample(
  df_groups, sample_key = "sample_id", group_key = "group",
  cluster_key = "cell_type", patient_key = "patient",
  unit = "patient",
  neighbors.k = 15, n_perms = 100, n_jobs = 1
)
```

# Inputs other than a data.frame

All `*_per_sample()` helpers accept three kinds of input:

* A `Seurat` object with multiple images. `sample_key` is a column of
  `meta.data` whose values match `names(obj@images)`.
* A `list` of `Seurat` objects, where each may itself have multiple
  images. `source_index` in `build_sample_design()` records the list
  origin.
* A `data.frame` with `x`, `y`, the cluster column, and the sample /
  group / (optional) patient columns.

`build_sample_design()` is the small helper that turns any of these into
the design table used downstream — call it directly if you only want the
sample / group / patient mapping.

```{r design}
build_sample_design(
  df_groups,
  sample_key  = "sample_id",
  group_key   = "group",
  patient_key = "patient"
)
```
