---
title: "Disease-group comparison with spatialCooccur"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Disease-group comparison with spatialCooccur}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.5,
  dpi = 90
)
```

This vignette walks through the disease- / condition-group comparison
workflow in **spatialCooccur**. The pattern is:

1. Collect per-sample spatial transcriptomics data (one image per sample,
   or multiple images grouped by patient).
2. Compute a spatial co-occurrence score *per sample* with one of the
   `*_per_sample()` helpers.
3. Test for differences between groups (e.g. disease vs. control) with
   `compare_groups()`.
4. Visualize with `plot_group_delta_heatmap()`, `plot_pair_boxplot()`, or
   `plot_volcano_groups()`.

```{r load}
library(spatialCooccur)
```

# Simulate a two-group dataset

`generate_sim_groups()` wraps `generate_sim()` to produce
`n_samples_per_group` samples per condition, each with a possibly noised
`close_ratio` around the group baseline. Here we use `test_type = "distribute"`
so that planted co-localization is at the spatial scale the local
neighborhood graph is sensitive to.

```{r simulate}
set.seed(1)
df_groups <- generate_sim_groups(
  n_samples_per_group = 4,
  group_close_ratio = list(disease = 0.85, control = 0.15),
  n_types = 5,
  n_cells = 400,
  test_type = "distribute",
  distance_param = 15,
  between_sample_noise = 0.03,
  seed = 1
)
head(df_groups)
table(df_groups$sample_id, df_groups$group)
```

`df_groups` is a long-format data.frame with columns `x`, `y`,
`cell_type`, `sample_id`, `group`, `patient`. The same shape is what the
`*_per_sample()` helpers accept directly; for Seurat input they read the
equivalent fields from `meta.data` and `@images`.

# Per-sample neighborhood enrichment

```{r per_sample, message = FALSE}
per_sample <- nhood_enrichment_per_sample(
  df_groups,
  sample_key   = "sample_id",
  group_key    = "group",
  cluster_key  = "cell_type",
  patient_key  = "patient",
  neighbors.k  = 15,
  n_perms      = 100,
  n_jobs       = 1,
  seed         = 1
)
head(per_sample)
```

Each row is `sample_id x cluster_i x cluster_j`. The factor levels of
`cell_type` are harmonized across samples so that the `cluster_i`,
`cluster_j` columns are directly comparable.

For the radius-based ratio, the local interaction score, and the
interaction-spot count, use the parallel helpers:

```{r other_per_sample, eval = FALSE}
cooccur_ratio_per_sample(
  df_groups, sample_key = "sample_id", group_key = "group",
  cluster_key = "cell_type", patient_key = "patient",
  radius = 30, k = 15
)

cooccur_local_per_sample(
  df_groups, sample_key = "sample_id", group_key = "group",
  cluster_key = "cell_type",
  cluster_x = "cell_type_1", cluster_y = "cell_type_2",
  patient_key = "patient",
  neighbors.k = 15, radius = 30
)
```

# Group comparison

`compare_groups()` runs a per-cluster-pair test between two groups. Set
`ref_group` to make the sign of `effect` interpretable
(`effect = mean(test) - mean(reference)`).

```{r wilcox}
cmp_wilcox <- compare_groups(
  per_sample,
  value     = "zscore",
  method    = "wilcox",
  adjust    = "BH",
  ref_group = "control"
)
head(cmp_wilcox)
```

When you have multiple images per patient, prefer a mixed model (random
intercept on patient) or a patient-blocked permutation:

```{r lmm, eval = requireNamespace("lme4", quietly = TRUE)}
cmp_lmm <- compare_groups(
  per_sample,
  value       = "zscore",
  method      = "lmm",
  patient_key = "patient",
  ref_group   = "control"
)
head(cmp_lmm)
```

```{r perm}
cmp_perm <- compare_groups(
  per_sample,
  value       = "zscore",
  method      = "perm",
  patient_key = "patient",
  n_perms     = 500,
  ref_group   = "control",
  seed        = 1
)
head(cmp_perm)
```

# Visualization

```{r plots, eval = requireNamespace("ggplot2", quietly = TRUE), fig.show = "hold"}
plot_group_delta_heatmap(cmp_wilcox)
plot_volcano_groups(cmp_wilcox, label_top = 5)
plot_pair_boxplot(
  per_sample,
  value     = "zscore",
  pairs     = data.frame(cluster_i = "cell_type_1",
                         cluster_j = "cell_type_2"),
  add_p     = TRUE,
  ref_group = "control"
)
```

# Patient-level aggregation

If you have multiple images per patient and want one observation per
patient, pass `unit = "patient"` to the `*_per_sample()` helpers. Each
image-level result is then averaged within patient before group
comparison:

```{r patient_level, eval = FALSE}
per_patient <- nhood_enrichment_per_sample(
  df_groups, sample_key = "sample_id", group_key = "group",
  cluster_key = "cell_type", patient_key = "patient",
  unit = "patient",
  neighbors.k = 15, n_perms = 100, n_jobs = 1
)
```

# Inputs other than a data.frame

All `*_per_sample()` helpers accept three kinds of input:

* A `Seurat` object with multiple images. `sample_key` is a column of
  `meta.data` whose values match `names(obj@images)`.
* A `list` of `Seurat` objects, where each may itself have multiple
  images. `source_index` in `build_sample_design()` records the list
  origin.
* A `data.frame` with `x`, `y`, the cluster column, and the sample /
  group / (optional) patient columns.

`build_sample_design()` is the small helper that turns any of these into
the design table used downstream — call it directly if you only want the
sample / group / patient mapping.

```{r design}
build_sample_design(
  df_groups,
  sample_key  = "sample_id",
  group_key   = "group",
  patient_key = "patient"
)
```
