---
title: Tutorial for spatial neighborhood analysis (SNA) & spatial co-localization score (sCLS) using 10X public data
author: 
  name: Jun Inamo
  email: jun.inamo@cuanschutz.edu
  affiliation: Computational Omics and Systems Immunology (COSI) Lab, Division of Rheumatology and Center for Health AI, University of Colorado School of Medicine, CO, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  BiocStyle::html_document:
    toc_float: true
---



```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
# BiocManager::install("BiocStyle")
library(BiocStyle)
```

```{r, message=FALSE, warning=FALSE, include=FALSE}

figure_dir="sim_cooccur"
dir.create(paste0("./output/figure/",figure_dir), showWarnings = FALSE)

```

```{r, message=FALSE, warning=FALSE}
library(Seurat)
library(magrittr)
library(dplyr)
# devtools::install_github("zhanghao-njmu/SCP")
library(SCP)
library(ggplot2)
library(circlize)
library(ComplexHeatmap)
library(gridExtra)

library(spatialCooccur)

BuildSNNSeurat <- function (data.use, k.param = 30, prune.SNN = 1/15, nn.eps = 0) {
  my.knn <- nn2(data = data.use, k = k.param, searchtype = "standard", eps = nn.eps)
  nn.ranked <- my.knn$nn.idx
  
  snn_res <- ComputeSNN(nn_ranked = nn.ranked, prune = prune.SNN)
  rownames(snn_res) <- row.names(data.use)
  colnames(snn_res) <- row.names(data.use)
  return(snn_res)
}
environment(BuildSNNSeurat) <- asNamespace("Seurat")
```

# Xenium Human Breast Gene Expression

- curl -O     https://cf.10xgenomics.com/samples/xenium/2.0.0/Xenium_V1_human_Breast_2fov/Xenium_V1_human_Breast_2fov_outs.zip

## Preprocessing & Cell annotation

```{r}
path <- "./../10X_Xenium_sample/Xenium_V1_human_Breast_2fov_outs/"
data_name = stringr::str_split(path, "/")[[1]][length(stringr::str_split(path, "/")[[1]])-1]
dir.create(paste0("./output/figure/",figure_dir,"/",data_name), showWarnings = FALSE)

# Load the Xenium data
data <- ReadXenium(path, outs = c("matrix", "microns"), type = c("centroids", "segmentations"))
## continue the regular LoadXenium
segmentations.data <- list(
  centroids = CreateCentroids(data$centroids),
  segmentation = CreateSegmentation(data$segmentations))
coords <- CreateFOV(
  coords = segmentations.data, 
  type = c("segmentation", "centroids"), 
  molecules = data$microns, 
  assay = "Spatial")
xenium.obj <- CreateSeuratObject(
  counts = data$matrix[["Gene Expression"]], 
  assay = "Spatial")
xenium.obj[["BlankCodeword"]] <- CreateAssayObject(counts = data$matrix[["Unassigned Codeword"]])
xenium.obj[["ControlCodeword"]] <- CreateAssayObject(counts = data$matrix[["Negative Control Codeword"]])
xenium.obj[["ControlProbe"]] <- CreateAssayObject(counts = data$matrix[["Negative Control Probe"]])
xenium.obj[["fov"]] <- coords
rm(data); gc(); gc()
rm(coords); gc(); gc()
rm(segmentations.data); gc(); gc()

xenium.obj <- subset(xenium.obj, subset = nCount_Spatial > 0)

print(dim(xenium.obj@assays$Spatial$counts))
xenium.obj@assays$Spatial$counts[1:5,1:5]
head(xenium.obj@meta.data)
summary(xenium.obj@meta.data$nCount_Spatial)
summary(xenium.obj@meta.data$nFeature_Spatial)

VlnPlot(xenium.obj, features = c("nFeature_Spatial", "nCount_Spatial"), ncol = 2, pt.size = 0)

ImageDimPlot(xenium.obj, fov = "fov", molecules = c("TUBB2B", "PELI1", "CENPF", "KRT23","PDGFRB","ITGAX","KRT14","GZMA"), nmols = 20000)

```


```{r}

xenium.obj <- SCTransform(xenium.obj, assay = "Spatial",
                          conserve.memory = TRUE, vst.flavor="v2")
xenium.obj <- RunPCA(xenium.obj, assay = "SCT", verbose = FALSE)
xenium.obj <- FindNeighbors(xenium.obj, 
                     reduction = "pca", 
                     k.param = 30,
                     dims = 1:30)
xenium.obj <- RunUMAP(xenium.obj, 
               reduction = "pca", 
               n.neighbors = 30L,
               min.dist = 0.3,
               dims = 1:30)

print("Clustering...")
# Louvain
snn_pcs <- BuildSNNSeurat(xenium.obj[["pca"]]@cell.embeddings[,1:30], 
                          nn.eps = 0)

resolution_list <- c(0.2, 0.4, 0.6, 0.8, 1.0)
ids_cos <- Reduce(cbind, parallel::mclapply(resolution_list, function(res_use) {
  Seurat:::RunModularityClustering(SNN = snn_pcs, 
                                   modularity = 1, 
                                   resolution = res_use, 
                                   algorithm = 3, 
                                   n.start = 10, 
                                   n.iter = 10, random.seed = 0, print.output = FALSE, 
                                   temp.file.location = NULL, edge.file.name = NULL)    
}, mc.cores = min(16, length(resolution_list))))
ids_cos %<>% data.frame()
colnames(ids_cos) <- sprintf("res_%.2f", resolution_list)

ids_cos$res_0.20 <- as.character(ids_cos$res_0.20)
ids_cos$res_0.40 <- as.character(ids_cos$res_0.40)
ids_cos$res_0.60 <- as.character(ids_cos$res_0.60)
ids_cos$res_0.80 <- as.character(ids_cos$res_0.80)
ids_cos$res_1.00 <- as.character(ids_cos$res_1.00)

rownames(ids_cos) = rownames(xenium.obj@meta.data)
ids_cos <- ids_cos %>%
  dplyr::mutate(across(everything(), ~ factor(.x, levels = sort(unique(.x)))))
xenium.obj <- AddMetaData(xenium.obj, ids_cos)
head(xenium.obj@meta.data)


resolution = "0.60"
Idents(xenium.obj) = xenium.obj@meta.data[,paste0("res_",resolution)]

cluster_col = paste0("res_",resolution)

```

```{r}

g = CellDimPlot(
  srt = xenium.obj, 
  group.by = cluster_col, 
  reduction = "UMAP", theme_use = "theme_blank",
  raster = FALSE,
  stat_plot_size = 3
) 
g 


```

```{r}
xenium.obj.markers <- FindAllMarkers(xenium.obj, only.pos = TRUE)
xenium.obj.markers %>%
  group_by(cluster) %>%
  dplyr::slice_max(avg_log2FC, n = 10) %>%
  as.data.frame() 
```


Cluster annotation:
0	Basal-like epithelial cells / Myoepithelial cells
1	Proliferating epithelial cells
2	Basal/myoepithelial tumor cells
3	Unclassified immune-like cells
4	Inflammatory immune cells (e.g. B cells / monocytes)
5	Cancer-associated fibroblasts (CAFs)
6	Macrophages (TAMs)
7	Cytotoxic / Activated T cells
8	Endothelial cells


```{r}
table(xenium.obj@meta.data[,cluster_col])
xenium.obj@meta.data$new_cluster = dplyr::case_when(
  xenium.obj@meta.data[,cluster_col] == "0" ~ "BasalMyo_epithelial_cells",
  xenium.obj@meta.data[,cluster_col] == "1" ~ "Proliferating_epithelial_cells",
  xenium.obj@meta.data[,cluster_col] == "2" ~ "BasalMyo_tumor_cells",
  xenium.obj@meta.data[,cluster_col] == "3" ~ "Unclassified_immune-like_cells",
  xenium.obj@meta.data[,cluster_col] == "4" ~ "Inflammatory_immune_cells",
  xenium.obj@meta.data[,cluster_col] == "5" ~ "Cancer-associated_fibroblasts",
  xenium.obj@meta.data[,cluster_col] == "6" ~ "Tumor-associated_macrophages",
  xenium.obj@meta.data[,cluster_col] == "7" ~ "Cytotoxic_T",
  xenium.obj@meta.data[,cluster_col] == "8" ~ "Endothelial_cells"
)
table(xenium.obj@meta.data$new_cluster)
cluster_col = "new_cluster"
Idents(xenium.obj) = xenium.obj@meta.data$new_cluster
```

```{r}

cluster_COLORS = manual_colors
names(cluster_COLORS) = levels(Idents(xenium.obj))
cluster_COLORS = na.omit(cluster_COLORS)
g = CellDimPlot(
  srt = xenium.obj, 
  group.by = cluster_col, 
  reduction = "UMAP", theme_use = "theme_blank",
  raster = FALSE,
  stat_plot_size = 1
) &
  scale_color_manual(values = cluster_COLORS) &
  guides(color = guide_legend(override.aes = list(size=5,
                                                 alpha = 1),
                             title = "cluster",
                             ncol = 1))
g 
ggsave(paste0("./output/figure/",figure_dir,"/",data_name,"/CellDimPlot.pdf"), g, width=7, height=4, dpi=300)
```

```{r}
g = ImageDimPlot(xenium.obj,
             fov = "fov", 
             size = 1.2,
             group.by = cluster_col, 
             dark.background = F)  &
  scale_fill_manual(values = cluster_COLORS) &
  guides(fill = guide_legend(override.aes = list(size=5,
                                                 alpha = 1),
                             title = "cluster",
                             ncol = 1))
g 
ggsave(paste0("./output/figure/",figure_dir,"/",data_name,"/ImageDimPlot_byClusters.pdf"), g, width=6, height=4, dpi=300)

print(xenium.obj)
```

## Spatial neighborhood analysis (SNA, cell type level analysis)

```{r}
n_perm = 100
neighbors.k_ = 30 # Number of neighbors to search
seed = 1234

start_time = Sys.time()
xenium.obj <- nhood_enrichment.Seurat(
  xenium.obj,
  cluster_key = cluster_col, 
  neighbors.k = neighbors.k_, 
  connectivity_key = "nn", 
  transformation = TRUE,
  n_perms = n_perm, seed = seed, n_jobs = 4
)
end_time = Sys.time()
print("Elapsed time:")
difftime(end_time, start_time, units = "secs")
```

```{r, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
mat = xenium.obj@misc[[paste0(cluster_col,"_nhood_enrichment")]]$zscore
colnames(mat) = gsub("^Cluster","",colnames(mat))
rownames(mat) = gsub("^Cluster","",rownames(mat))

pval_mat <- 1 - pnorm(mat)

fdr_vec <- p.adjust(as.vector(pval_mat), method = "BH")
fdr_mat <- matrix(fdr_vec, nrow=nrow(mat), ncol=ncol(mat),
                  dimnames = dimnames(mat))
sig_mat <- ifelse(fdr_mat < 0.05, "**", ifelse(fdr_mat > 0.05 & fdr_mat < 0.1, "*", ""))

common_names <- intersect(rownames(mat), colnames(mat))
for (nm in common_names) {
  mat[nm, nm] <- NA
  sig_mat[nm, nm] <- ""
}

heatmap <- Heatmap(mat,
                   name = "Z-score",
                   col = colorRamp2(c(-2, 0, 2), c("#0072B5FF", "white", "#BC3C29FF")), 
                   show_row_names = TRUE, 
                   show_column_names = TRUE,  
                   cluster_rows = TRUE,  
                   cluster_columns = TRUE,  
                   #show_column_dend = FALSE,
                   #show_row_dend = FALSE,
                   row_title = "",  
                   column_title = paste0("Spatial Neigborhood Enrichment\n",data_name),
                   rect_gp = gpar(col = "black", lwd = 0.3),
                   na_col = "black",         
                   column_names_gp = grid::gpar(fontsize = 10),
                   row_names_gp = grid::gpar(fontsize = 10),

                   cell_fun = function(j, i, x, y, width, height, fill) {
                     if(sig_mat[i, j] == "**") {
                       grid.text("**", 
                                 x = x,
                                 y = y - 0.2 * height, 
                                 gp = gpar(fontsize = 15, col = "white", fontface = "bold"))
                     }
                     if(sig_mat[i, j] == "*") {
                       grid.text("*", 
                                 x = x,
                                 y = y - 0.2 * height,  
                                 gp = gpar(fontsize = 15, col = "white", fontface = "bold"))
                     }
                   }
)

#pdf(paste0("./output/figure/",figure_dir,"/",data_name,"/SNA_heatmap.pdf"), width=4.5, height=5.5)
draw(heatmap, 
     merge_legend = TRUE,
     heatmap_legend_side = "bottom", 
     annotation_legend_side = "bottom")
#dev.off()
```

## Spatial co-localization score (sCLA, cell-cell level analysis)


```{r, message=FALSE, warning=FALSE}

cluster_x <- "Cytotoxic_T"
cluster_y <- "Cancer-associated_fibroblasts"
table(Idents(xenium.obj))
radius_ = 30 # Radius to search for neighbors (µm, cell_type_2) around anchor cells (cell_type_1)

start_time = Sys.time()
cooccur_local_df <- cooccur_local.Seurat(
  xenium.obj,
  cluster_key      = cluster_col,
  sample_key       = NULL,
  cluster_x        = cluster_x,
  cluster_y        = cluster_y,
  connectivity_key = "nn",
  neighbors.k      = neighbors.k_, 
  radius           = radius_,
  maxnsteps        = 15
)
end_time = Sys.time()
print("Elapsed time:")
difftime(end_time, start_time, units = "secs")
summary(cooccur_local_df)
```

```{r, message=FALSE, warning=FALSE}
xenium.obj = AddMetaData(xenium.obj, cooccur_local_df)

g = ImageFeaturePlot(xenium.obj, 
                     features = paste0("cooccur_local_",cluster_x,"_",cluster_y), 
                     #dark.background = F,
                     cols = c("white", "red"))
g
ggsave(paste0("./output/figure/",figure_dir,"/",data_name,"/ImageFeaturePlot_",cluster_x,"_",cluster_y,".pdf"), g, width=6, height=4, dpi=300)

```


<br/><br/>


# Xenium Mouse Brain

- wget https://cf.10xgenomics.com/samples/xenium/1.0.2/Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP/Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP_outs.zip
- unzip Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP_outs.zip
  
## Preprocessing & Cell annotation

```{r}
path <- "./../10X_Xenium_sample/Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP_outs/"
data_name = stringr::str_split(path, "/")[[1]][length(stringr::str_split(path, "/")[[1]])-1]
dir.create(paste0("./output/figure/",figure_dir,"/",data_name), showWarnings = FALSE)

xenium.obj <- LoadXenium(path, fov = "fov")
# remove cells with 0 counts
xenium.obj <- subset(xenium.obj, subset = nCount_Xenium > 0)

VlnPlot(xenium.obj, features = c("nFeature_Xenium", "nCount_Xenium"), ncol = 2, pt.size = 0)


```
```{r, fig.width=8, fig.height=8}
ImageDimPlot(xenium.obj, 
             fov = "fov", 
             size = 0.5,
             molecules = c("Gad1", "Sst", "Pvalb", "Gfap"), nmols = 20000)
```


```{r}
xenium.obj <- SCTransform(xenium.obj, assay = "Xenium")
xenium.obj <- RunPCA(xenium.obj, npcs = 30, features = rownames(xenium.obj))
xenium.obj <- RunUMAP(xenium.obj, dims = 1:30)
xenium.obj <- FindNeighbors(xenium.obj, reduction = "pca", dims = 1:30)
xenium.obj <- FindClusters(xenium.obj, resolution = 0.3)

```

```{r}

cluster_col="seurat_clusters"
g = CellDimPlot(
  srt = xenium.obj, 
  group.by = cluster_col, 
  reduction = "UMAP", theme_use = "theme_blank",
  raster = FALSE,
  stat_plot_size = 3
) 
g 


```

```{r, fig.width=8, fig.height=12}
FeaturePlot(xenium.obj, features = c("Cux2", "Bcl11b", "Foxp2", "Gad1", "Sst", "Gfap"))
```

```{r, fig.width=8, fig.height=8}
ImageDimPlot(xenium.obj, cols = "polychrome", 
             size = 0.75)
```

```{r}
# devtools::install_github("dmcable/spacexr", build_vignettes = FALSE)
library(spacexr)

query.counts <- GetAssayData(xenium.obj, assay = "Xenium", slot = "counts")
coords <- GetTissueCoordinates(xenium.obj[["fov"]], which = "centroids")
rownames(coords) <- coords$cell
coords$cell <- NULL
query <- SpatialRNA(coords, query.counts, colSums(query.counts))
```

```{r, message=FALSE, warning=FALSE}
# allen.corted.ref can be downloaded here:
# https://www.dropbox.com/s/cuowvm4vrf65pvq/allen_cortex.rds?dl=1
allen.cortex.ref <- readRDS("./../10X_Xenium_sample/allen_cortex.rds")
allen.cortex.ref <- UpdateSeuratObject(allen.cortex.ref)

Idents(allen.cortex.ref) <- "subclass"
# remove CR cells because there aren't enough of them for annotation
allen.cortex.ref <- subset(allen.cortex.ref, subset = subclass != "CR")
counts <- GetAssayData(allen.cortex.ref, assay = "RNA", slot = "counts")
cluster <- as.factor(allen.cortex.ref$subclass)
names(cluster) <- colnames(allen.cortex.ref)
nUMI <- allen.cortex.ref$nCount_RNA
names(nUMI) <- colnames(allen.cortex.ref)
nUMI <- colSums(counts)
levels(cluster) <- gsub("/", "-", levels(cluster))
reference <- Reference(counts, cluster, nUMI)
```

```{r, message=FALSE, warning=FALSE}
# run RCTD with many cores
RCTD <- create.RCTD(query, reference, max_cores = 8)
RCTD <- run.RCTD(RCTD, doublet_mode = "doublet")
```

```{r}
annotations.df <- RCTD@results$results_df
annotations <- annotations.df$first_type
names(annotations) <- rownames(annotations.df)
xenium.obj$predicted.celltype <- annotations
keep.cells <- Cells(xenium.obj)[!is.na(xenium.obj$predicted.celltype)]
xenium.obj <- subset(xenium.obj, cells = keep.cells)
```

```{r}
cluster_col="predicted.celltype"
Idents(xenium.obj) = xenium.obj@meta.data[,cluster_col]

```

```{r}
markers = xenium.obj.markers %>%
  group_by(cluster) %>%
  dplyr::slice_max(avg_log2FC, n = 1) %>%
  as.data.frame() %>%
  pull(gene) %>%
  unique()
marker_COLORS = manual_colors
names(marker_COLORS) = markers
  
ImageDimPlot(xenium.obj, fov = "fov", molecules = markers, nmols = 20000) &
  scale_color_manual(values = marker_COLORS) 
```


```{r}

cluster_COLORS = manual_colors
names(cluster_COLORS) = levels(Idents(xenium.obj))
cluster_COLORS = na.omit(cluster_COLORS)
g = CellDimPlot(
  srt = xenium.obj, 
  group.by = cluster_col, 
  reduction = "UMAP", theme_use = "theme_blank",
  raster = FALSE,
  stat_plot_size = 1
) &
  scale_color_manual(values = cluster_COLORS) &
  guides(color = guide_legend(override.aes = list(size=5,
                                                 alpha = 1),
                             title = "cluster",
                             ncol = 1))
g 
ggsave(paste0("./output/figure/",figure_dir,"/",data_name,"/CellDimPlot.pdf"), g, width=7, height=4, dpi=300)

print(xenium.obj)
```

```{r}
g = ImageDimPlot(xenium.obj,
             fov = "fov", 
             size = 0.5,
             group.by = cluster_col)  &
  scale_fill_manual(values = cluster_COLORS)
g 
ggsave(paste0("./output/figure/",figure_dir,"/",data_name,"/ImageDimPlot_byClusters.pdf"), g, width=6, height=4, dpi=300)
```

## Spatial neighborhood analysis (SNA, cell type level analysis)

```{r}
n_perm = 100
neighbors.k_ = 30 # Number of neighbors to search
seed = 1234

start_time = Sys.time()
xenium.obj <- nhood_enrichment.Seurat(
  xenium.obj,
  cluster_key = cluster_col, 
  neighbors.k = neighbors.k_, 
  connectivity_key = "nn", 
  transformation = TRUE,
  n_perms = n_perm, seed = seed, n_jobs = 4
)
end_time = Sys.time()
print("Elapsed time:")
difftime(end_time, start_time, units = "secs")
```

```{r, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
mat = xenium.obj@misc[[paste0(cluster_col,"_nhood_enrichment")]]$zscore
colnames(mat) = gsub("^Cluster","",colnames(mat))
rownames(mat) = gsub("^Cluster","",rownames(mat))

pval_mat <- 1 - pnorm(mat)

fdr_vec <- p.adjust(as.vector(pval_mat), method = "BH")
fdr_mat <- matrix(fdr_vec, nrow=nrow(mat), ncol=ncol(mat),
                  dimnames = dimnames(mat))

sig_mat <- ifelse(fdr_mat < 0.05, "**", ifelse(fdr_mat > 0.05 & fdr_mat < 0.1, "*", ""))

common_names <- intersect(rownames(mat), colnames(mat))
for (nm in common_names) {
  mat[nm, nm] <- NA
  sig_mat[nm, nm] <- ""
}

heatmap <- Heatmap(mat,
                   name = "Z-score",
                   col = colorRamp2(c(-2, 0, 2), c("#0072B5FF", "white", "#BC3C29FF")), 
                   show_row_names = TRUE, 
                   show_column_names = TRUE,  
                   cluster_rows = TRUE,  
                   cluster_columns = TRUE,  
                   #show_column_dend = FALSE,
                   #show_row_dend = FALSE,
                   row_title = "",  
                   column_title = paste0("Spatial Neigborhood Enrichment\n",data_name),
                   rect_gp = gpar(col = "black", lwd = 0.3),
                   na_col = "black",         
                   column_names_gp = grid::gpar(fontsize = 10),
                   row_names_gp = grid::gpar(fontsize = 10),

                   cell_fun = function(j, i, x, y, width, height, fill) {
                     if(sig_mat[i, j] == "**") {
                       grid.text("**", 
                                 x = x,
                                 y = y - 0.2 * height,  
                                 gp = gpar(fontsize = 15, col = "white", fontface = "bold"))
                     }
                     if(sig_mat[i, j] == "*") {
                       grid.text("*", 
                                 x = x,
                                 y = y - 0.2 * height, 
                                 gp = gpar(fontsize = 15, col = "white", fontface = "bold"))
                     }
                   }
)

#pdf(paste0("./output/figure/",figure_dir,"/",data_name,"/SNA_heatmap.pdf"), width=4.5, height=5.5)
draw(heatmap, 
     merge_legend = TRUE,
     heatmap_legend_side = "bottom", 
     annotation_legend_side = "bottom")
#dev.off()
```

## Spatial co-localization score (sCLA, cell-cell level analysis)


```{r, message=FALSE, warning=FALSE}

cluster_x <- "L6 IT"
cluster_y <- "L6 CT"
table(Idents(xenium.obj))
radius_ = 30 # Radius to search for neighbors (µm, cell_type_2) around anchor cells (cell_type_1)

start_time = Sys.time()
cooccur_local_df <- cooccur_local.Seurat(
  xenium.obj,
  cluster_key      = cluster_col,
  sample_key       = NULL,
  cluster_x        = cluster_x,
  cluster_y        = cluster_y,
  connectivity_key = "nn",
  neighbors.k      = neighbors.k_, 
  radius           = radius_,
  maxnsteps        = 15
)
end_time = Sys.time()
print("Elapsed time:")
difftime(end_time, start_time, units = "secs")


colnames(cooccur_local_df) = gsub(" ", "_", colnames(cooccur_local_df))
summary(cooccur_local_df)
```

```{r, message=FALSE, warning=FALSE}
xenium.obj = AddMetaData(xenium.obj, cooccur_local_df)

g = ImageFeaturePlot(xenium.obj, 
                     features = gsub(" ", "_", paste0("cooccur_local_",cluster_x,"_",cluster_y)), 
                     #dark.background = F,
                     cols = c("white", "red"))
g
ggsave(paste0("./output/figure/",figure_dir,"/",data_name,"/ImageFeaturePlot_",gsub(" ", "_", cluster_x),"_",gsub(" ", "_", cluster_y),".pdf"), g, width=6, height=4, dpi=300)

```

```{r, fig.width=12, fig.height=8}

cluster_COLORS_ = cluster_COLORS[cluster_x]
g_x = ImageDimPlot(xenium.obj,
             fov = "fov", 
             size = 0.3,
             group.by = cluster_col)  &
  scale_fill_manual(values = cluster_COLORS_) &
  guides(fill = guide_legend(override.aes = list(size=5,
                                                 alpha = 1),
                             title = "cluster",
                             ncol = 1))
cluster_COLORS_ = cluster_COLORS[cluster_y]
g_y = ImageDimPlot(xenium.obj,
             fov = "fov", 
             size = 0.3,
             group.by = cluster_col)  &
  scale_fill_manual(values = cluster_COLORS_) &
  guides(fill = guide_legend(override.aes = list(size=5,
                                                 alpha = 1),
                             title = "cluster",
                             ncol = 1))
g_x + g_y
ggsave(paste0("./output/figure/",figure_dir,"/",data_name,"/ImageDimPlot_",gsub(" ", "_", cluster_x),"_",gsub(" ", "_", cluster_y),".pdf"), g_x + g_y, width=6, height=4, dpi=300)

```

<br/><br/>

***


# sessionInfo

```{r}
sessionInfo()
```


